Placing the spotted T Tauri star LkCa 4 on an HR diagram

### 2017-01-06 document/submission/ms.tex
Big caveats about comparing to non-spotted models.

Ages and masses of young stars are often estimated by comparing their observed luminosities and effective temperatures to pre-main sequence stellar evolutionary model tracks, but magnetic fields and starspots complicate this process. In order to understand their influence, we study in detail the heavily-spotted weak-lined \emph{T-Tauri} star LkCa 4 by searching for spectral signatures of radiation originating from the starspot or starspot groups.  We introduce a new methodology for constraining both the starspot filling factor and the typical spot temperature by fitting two-temperature stellar atmosphere models constructed from \PHOENIX\ synthetic spectra to a high-resolution near-IR IGRINS spectrum.  Clearly discernable spectral features arise from both a hot photospheric component $\teffa\sim4100$ K and to a cool component $\teffb\sim2700-3000$ K, with a best-fit filling factor of $\sim80$\%  for the cool photosphere at the time of the observation.  These results are supported by comparisons of this mix of hot and cool photospheric temperatures by analyses of the spectral energy distribution, rotational modulation of colors and of TiO band strengths, and low-resolution optical/near-IR spectroscopy.  LkCa 4 has either a warm photosphere mostly covered in cool spots or is a very cool star with a small hot spot.  The revised effective temperature and luminosity make LkCa 4 appear much younger and lower mass when improperly compared to previous estimates and with unspotted model tracks, motivating the production and adoption of spotted model tracks.  Biases from starspots likely afflict most fully convective young stars and contribute to uncertainties in open cluster ages inferred from low-mass stars and in large apparent age spreads of those clusters. In some spectral regions starspots act as a featureless ``veiling'' continuum owing to the large rotational broadening ($\vsini\sim28$ km~s$^{-1}$) and heavy line-blanketing in cool star spectra.  Evidence for anti-correlations in velocity of the the warm and cool components is found.
\end{abstract}

@@

Starspot line depth ratio analysis has traditionally been limited to isolated portions of spectrum that possess absorption lines attributable only to cool photospheric components and spectral lines arising from the warmer ambient photosphere \citep[\emph{e.g.}][]{neff95, oneal01}.  The apparent veiling of the lines and their respective temperature dependences can be combined to solve for the cool photosphere and ambient photosphere temperatures and relative areal coverages.  This line depth ratio method suffers from the need to identify portions of the spectrum that possess easily identifiable strong lines, and from the need to assemble large atlases of observed spotless spectral templates to which the spotted star spectrum can be compared.  In this section, we introduce a generalization of the line depth ratio analysis which employs pixel-by-pixel modeling of all spectral orders.  By using all the spectral data, this strategy has the power to constrain starspot properties at relatively low filling factors and to identify weak lines that originate from the cool photosphere.

We took two approaches to characterizing the photospheric temperature from a spotted-star spectrum.  First, 
using the spectral fitting approach introduced in Section 3.1 \ref{sec:methods}, we separately fit the optical (ESPaDOnS) and near-IR (IGRINS) spectra with distinct, single photospheric component models; these fits are described in Sections \ref{sec:ESP_starfish} and \ref{sec:IGR_starfish} respectively, and then compared in \ref{sec:whyNearIR}.
Second, we use a two temperature mixture model to perform a fit to the near-IR spectra in Section \ref{sec:two_tempIGRINS}.  The near-IR spectral range is preferred over the optical, since the cool photosphere emits most of its flux in the near-IR, enhancing the likelihood of direct detection of emission from the cool component.

@@

We performed spectral fitting on an ESPaDOnS spectrum acquired on 2014 January 11 (JD 2456668.9).  The spectrum was split into subsets of $N=26$ chunks, corresponding approximately to spectral order boundaries.  Fitting was performed separately on each spectral chunk.  The spectral emulator was trained on stellar parameters in the range $\logg \in [3.5, 4.0]$, $\Z \in [-0.5, 0.5]$, and photospheric temperatures $\in [3500, 4200]$ K.

@@

We selected a subset of 32 of the 54 available IGRINS spectral orders\footnote{The prohibitive computational cost limited us from running all the spectral orders.} with low telluric absorption artifacts.  We performed unique spectral fitting on each order, with free stellar parameters $\vt = (\teff, \logg, \Z)$.  We used the same analysis procedure as described for the ESPaDOnS spectra, with one exception.  For the IGRINS $K-$band, we employed an expanded search range for the photosperic temperature, $\teff \in [3000, 4200]$ K.

We find a larger dispersion in the point estimates for the stellar parameters derived from the IGRINS data than those derived in the optical spectrum.  The most conspicuous trend is in the derived photospheric temperature as a function of wavelength shown in Figure \ref{fig:SingleTeffvsOrder}.  The photospheric temperature peaks at values of $\sim4200$~K in the short wavelength end of $H-$band and saturates at $<3500$ K at the long wavelength end of $H-$band.  The $K-$band shows even lower derived photospheric temperatures of $\sim3300$ K, or 700 K cooler than estimated from optical.  No single temperature can describe all the spectral lines present in the high resolution optical and IR spectra.  
Sources with such discrepancies have been seen previously, for example in Figures 4 and 5 in \citet{bouvier92}.  These discrepancies are circumstantial evidence for the detection of spectral features attributable to a second photospheric component.

@@

Some care should be taken when directly comparing results between the ESPaDOnS and IGRINS spectra since they were not taken at the same time.  The particular ESPaDOnS spectrum used in Section \ref{sec:ESP_starfish} was obtained when LkCa 4 ahd an estimated $V-$band brightness of 12.94 mag, compared with 12.84 mag at the time of the IGRINS spectrum.  The fainter magnitude during the ESPaDOnS spectrum acquisition implies a greater coverage fraction than during the IGRINS spectrum.  
The $\teff$ derived in the optical bands and short-wavelength end of $H$-band yield similar values of $\sim 4100$ K, suggesting the \emph{bulk}\footnote{The bulk appearance and disappearance of spectral lines is mostly controlled by the temperature of the emitting region of the local photosphere.  In other words, most of the variance in a spectrum is attributable to temperature, assuming near-solar metallicity. 
Starspots imbue dearths of flux in the line profiles of optical spectra, but these are secondary to the mere presence of temperature-sensitive lines.} spectral features are broadly consistent with emission from a single temperature component. A single ESPaDOnS order surrounding the TiO bands shows an exceptionally low estimated $\teff$.  The long wavelength portion of $H-$band and all of $K-$band are more sensitive to starspot spectral signatures than the shorter wavelength portions.

@@

Full-spectrum fitting was performed for 48 of the 54 available IGRINS spectral orders, omitting only the orders with the most pathological telluric spectral artifacts. We applied the mixture model as described in Section \ref{sec:methods} and Appendix \ref{methods-details}.  Standard MCMC convergence tests were evaluated.  The stellar parameter ranges were $\logg \in [3.5, 4.0]$ and $\Z \in [-0.5, 0.5]$; the \PHOENIX\ model spectrum temperature range\footnote{The \PHOENIX\ model spectra are indexed by $\teff$, but this term carries a different meaning for two-component photospheres.} was $\teffa, \teffb \in [2700, 4500]$.  $H-$band fits had normal distribution priors in place: solar metallicity to $\pm0.05$ dex, $logg=3.8\pm0.1$, and $\vsini=29\pm5$ km/s.

@@

Figure~\ref{fig:hrdiag} and Table~4 shows that this effective temperature is much cooler than previous measurements from optical spectra, which are clustered around 4000 K.  The luminosity does not change significantly because the broadband SED yields similar luminosities.
Since the Hayashi tracks are nearly vertical at this temperature and age, the range in effective temperatures corresponds to a mass range from $\sim 0.15-0.3$ $M_\odot$ from standard evolutionary tracks \citep{baraffe15}.

@@

The improved characterization of the photospheric emission and radius of LkCa 4 should lead to more accurate estimates of mass and age from pre-main sequence tracks.  However, our effective temperature and luminosity yield a mass of $0.2-0.3$ $M_\odot$ and an age of $\sim 0.5$ Myr in the \citet{baraffe15} evolutionary models\footnote{The \citet{baraffe15} evolutionary tracks were calculated using BT-Settl photospheric spectra \citep{allard14}, which may introduce some uncertainties when compared to results from the PHOENIX models used here.}.  This age is uncomfortably young and this mass is uncomfortably low.  LkCa 4 is not located deeply within a molecular cloud and is not directly associated with any nearby Class 0/1 stars, which would be expected for a 0.5 Myr star.  Dynamical measurements of masses of young stars with similar optical properties suggest that LkCa 4 should have a mass of $\sim 0.7-1.0$ M$_\odot$ \citep[e.g.][]{guilloteau14,czekala16,rizzuto16}.

Even though LkCa 4 is only a single data point, the inferred age and low mass may be evidence that strong magnetic fields are inhibiting convection in LkCa 4.  As a result, the surface is cooler and releases less radiation, slowing the contraction rate.  LkCa 4 then appears more luminous and cooler than expected for a star of its genuine age from evolutionary tracks that do not consider magnetic fields.   This shift to lower temperature may also help to resolve some of the age discrepancies between intermediate mass and low mass young stars \citep[e.g.][]{herczeg15}.  The corrected placement of spotted stars on the HR diagram reflects the accurate luminosity, radius, and effective temperature, but evolutionary models are not yet equipped to interpret such heavily spotted stars \citep{somers15}.

>>>>

Ages and masses of young stars are often estimated by comparing their observed luminosities and effective temperatures to pre-main sequence stellar evolutionary model tracks, but magnetic fields and starspots complicate this process. In order to understand their influence, we study in detail the heavily-spotted weak-lined \emph{T-Tauri} star LkCa 4 by searching for spectral signatures of radiation originating from the starspot or starspot groups.  We introduce a new methodology for constraining both the starspot filling factor and the typical spot temperature by fitting two-temperature stellar atmosphere models constructed from \PHOENIX\ synthetic spectra to a high-resolution near-IR IGRINS spectrum.  Clearly discernable spectral features arise from both a hot photospheric component $\teffa\sim4100$ K and to a cool component $\teffb\sim2700-3000$ K, with a best-fit filling factor of $\sim80$\%  for the cool photosphere at the time of the observation.  These results are supported by comparisons of this mix of hot and cool photospheric temperatures by analyses of the spectral energy distribution, rotational modulation of colors and of TiO band strengths, and low-resolution optical/near-IR spectroscopy.  LkCa 4 has either a warm photosphere mostly covered in cool spots or is a very cool star with a small hot spot.  The revised effective temperature and luminosity make LkCa 4 appear much younger and lower mass when compared to previous estimates with the same unspotted stellar evolution models, motivating the production and adoption of spotted evolutionary models.  Biases from starspots likely afflict most fully convective young stars and contribute to uncertainties in open cluster ages inferred from low-mass stars and in large apparent age spreads of those clusters.  In some spectral regions starspots act as a featureless ``veiling'' continuum owing to the large rotational broadening ($\vsini\sim28$ km~s$^{-1}$) and heavy line-blanketing in cool star spectra.  Evidence for anti-correlations in velocity of the the warm and cool components is found.
\end{abstract}

@@

Starspot line depth ratio analysis has traditionally been limited to isolated portions of spectrum that possess absorption lines attributable only to cool photospheric components and spectral lines arising from the warmer ambient photosphere \citep[\emph{e.g.}][]{neff95, oneal01}.  The apparent veiling of the lines and their respective temperature dependences can be combined to solve for the cool photosphere and ambient photosphere temperatures and relative areal coverages.  This line depth ratio method suffers from the need to identify portions of the spectrum that possess easily identifiable strong lines, and from the need to assemble large atlases of observed spotless spectral templates to which the spotted star spectrum can be compared.  In this section, we introduce a generalization of the line depth ratio analysis which employs pixel-by-pixel modeling of all echelle spectral orders.  By using all the spectral data, this strategy has the power to constrain starspot properties at relatively low filling factors and to identify weak lines that originate from the cool photosphere.

We took two approaches to characterizing the photospheric temperature from a spotted-star spectrum.  First, 
using the spectral fitting approach introduced in Section \ref{sec:methods}, we separately fit the optical (ESPaDOnS) and near-IR (IGRINS) spectra with distinct, single photospheric component models; these fits are described in Sections \ref{sec:ESP_starfish} and \ref{sec:IGR_starfish} respectively, and then compared in \ref{sec:whyNearIR}.
Second, we use a two temperature mixture model to perform a fit to the near-IR spectra in Section \ref{sec:two_tempIGRINS}.  The near-IR spectral range is preferred over the optical, since the cool photosphere emits most of its flux in the near-IR, enhancing the likelihood of direct detection of emission from the cool component.

@@

We performed spectral fitting on an ESPaDOnS spectrum acquired on 2014 January 11 (JD 2456668.9).  The spectrum was split into subsets of $N=26$ chunks, corresponding approximately to spectral order boundaries.  Fitting was performed separately on each spectral chunk.  The spectral emulator was trained on stellar parameters in the range $\logg \in [3.5, 4.0]$, $\Z \in [-0.5, 0.5]$, and photospheric temperatures $ T \in [3500, 4200]$ K.

@@

We selected a subset of 32 of the 54 available IGRINS spectral orders\footnote{The prohibitive computational cost limited us from running all the spectral orders.} with low telluric absorption artifacts.  We performed unique spectral fitting on each order, with free stellar parameters $\vt = (\teff, \logg, \Z)$.  We used the same analysis procedure as described for the ESPaDOnS spectra, with one exception.  For the IGRINS $K-$band, we employed an expanded search range for the photosperic temperature, $T \in [3000, 4200]$ K.

We find a larger dispersion in the point estimates for the stellar parameters derived from the IGRINS data than those derived in the optical spectrum.  The most conspicuous trend is in the derived photospheric temperature as a function of wavelength shown in Figure \ref{fig:SingleTeffvsOrder}.  The photospheric temperature peaks at values of $\sim4200$~K in the short wavelength end of $H-$band and saturates at $<3500$ K at the long wavelength end of $H-$band.  The $K-$band shows even lower derived photospheric temperatures of $\sim3300$ K, or 700 K cooler than estimated from optical.  No single photospheric temperature can describe all the spectral lines present in the high resolution optical and IR spectra.  
Sources with such discrepancies have been seen previously, for example in Figures 4 and 5 in \citet{bouvier92}.  These discrepancies are circumstantial evidence for the detection of spectral features attributable to a second photospheric component.

@@

Some care should be taken when directly comparing results between the ESPaDOnS and IGRINS spectra since they were not taken at the same time.  The particular ESPaDOnS spectrum used in Section \ref{sec:ESP_starfish} was obtained when LkCa 4 had an estimated $V-$band brightness of 12.94 mag, compared with 12.84 mag at the time of the IGRINS spectrum.  The fainter magnitude during the ESPaDOnS spectrum acquisition implies a greater coverage fraction of cool material than during the IGRINS spectrum acquisition.  
The photospheric temperature derived in the optical bands and short-wavelength end of $H$-band yield similar values of $\sim 4100$ K, suggesting the \emph{bulk}\footnote{The bulk appearance and disappearance of spectral lines is mostly controlled by the temperature of the emitting region of the local photosphere.  In other words, most of the variance in a spectrum is attributable to temperature, assuming near-solar metallicity. 
Starspots imbue dearths of flux in the line profiles of optical spectra, but these are secondary to the mere presence of temperature-sensitive lines.} spectral features are broadly consistent with emission from a single temperature component. A single ESPaDOnS order surrounding the TiO bands shows an exceptionally low estimated $\teff$.  The long wavelength portion of $H-$band and all of $K-$band are more sensitive to starspot spectral signatures than the shorter wavelength portions.

@@

Full-spectrum fitting was performed for 48 of the 54 available IGRINS spectral orders, omitting only the orders with the most pathological telluric spectral artifacts. We applied the mixture model as described in Section \ref{sec:methods} and Appendix \ref{methods-details}.  Standard MCMC convergence tests were evaluated.  The stellar parameter ranges were $\logg \in [3.5, 4.0]$ and $\Z \in [-0.5, 0.5]$; the \PHOENIX\ model spectrum temperature range was $\teffa, \teffb \in [2700, 4500]$.  $H-$band fits had normal distribution priors in place: solar metallicity to $\pm0.05$ dex, $logg=3.8\pm0.1$, and $\vsini=29\pm5$ km/s.

@@

Figure~\ref{fig:hrdiag} and Table~4 shows that this effective temperature is much cooler than previous measurements from optical spectra, which are clustered around 4000 K.  The luminosity does not change significantly because the broadband SED yields similar luminosities.
Since the Hayashi tracks are nearly vertical at this temperature and age, the range in effective temperatures corresponds to a mass range from $\sim 0.15-0.3$ $M_\odot$ from standard, \emph{unspotted} evolutionary tracks \citep{baraffe15}.

@@

In principle, the improved characterization of the photospheric emission and radius of LkCa 4 should lead to more accurate estimates of mass and age from pre-main sequence tracks.  However, most standard pre-main sequence stellar evolution models do not take into account the effects of starspots, and those that do \citep{somers15} limit cool spot areal coverage fractions to less than 50\%.  Comparison of LkCa 4 to \emph{unspotted} models is inherently flawed, but for the purposes of gauging the effect size of spots on derived parameters, our effective temperature and luminosity yield a mass of $0.2-0.3$ $M_\odot$ and an age of $\sim 0.5$ Myr in the \citet{baraffe15} evolutionary models\footnote{The \citet{baraffe15} evolutionary tracks were calculated using BT-Settl photospheric spectra \citep{allard14}, which may introduce minor uncertainties when compared to results from the PHOENIX models used here.}.  This age is much younger and this mass is much lower than previous estimates made assuming the warm component solely accounted for the effective temperature.

Even though LkCa 4 is only a single data point, the cool spot coverage fraction may be evidence that strong magnetic fields are inhibiting convection in LkCa 4.  As a result, the surface is cooler and releases less radiation, slowing the contraction rate.  LkCa 4 then appears more luminous and cooler than expected for a star of its genuine age from evolutionary tracks that do not consider magnetic fields.   This shift to lower temperature may also help to resolve some of the age discrepancies between intermediate mass and low mass young stars \citep[e.g.][]{herczeg15}.  The corrected placement of spotted stars on the HR diagram reflects the accurate luminosity, radius, and effective temperature, but evolutionary models are not yet equipped to interpret such heavily spotted stars \citep{somers15}.  In an HR digram, spotted evolutionary model tracks at a given mass and age shift to higher luminosity and lower effective temperature.  Such a \emph{spot-corrected} HR diagram position would alter the age and mass estimates to $> 0.2-0.3$ $M_\odot$ and $\> 0.5$ Myr; in other words, back in the direction of previous na\"{i}ve estimates.  However the full magnitude of the correction is an out-standing question.

%LkCa 4 is not located deeply within a molecular cloud and is not directly associated with any nearby Class 0/1 stars, which would be expected for a 0.5 Myr star.  Dynamical measurements of masses of young stars with similar optical properties suggest that LkCa 4 should have a mass of $\sim 0.7-1.0$ M$_\odot$ \citep[e.g.][]{guilloteau14,czekala16,rizzuto16}.
